### 2013-09-13 thesis/src/chapters/chapter-2.tex
refactored table

While scalability is a widely used term, it still lacks a clear and concise definition. Over the time, there were a few attempts to define it, yet not all of them were claimed as successful \cite{Hi90} \cite{DuRoWi06}. Hence, it is necessary to clarify this term before going into further discussion. Instinctively, scalability is perceived as ability of a system to accommodate an increasing number of elements or objects to process. In particular, we can point out different types of scalability that are affected by increased number of requests: \cite{Bo00}:
\begin{itemize}
	\item \textit{load scalability} - ability to work without delays and unproductive resource consumption at light, moderate, or heavy loads while making good use of available resources. Factors that may hinder load scalability include: scheduling shared resource, self-expansion, inadequate exploitation of parallelism
	\item \textit{space scalability} - memory requirements do not grow to intolerable levels as the number of items system supports increases
	\item \textit{space-time scalability} - system continues to function gracefully as the number of objects it encompasses increases by orders of magnitude
	\item \textit{structural scalability} - implementation or standards do not impede the growth of the number of objects system encompasses
\end{itemize}
Although, all of the aforementioned aspects are crucial for functioning of any healthy application, our work focuses only on the first type of scalability. The reasoning behind this statement is that, while all of these scalability types lies in direct responsibility of an application developer, the load scalability can be additionally improved by adding additional resources to a system. This brings us to a question what kind of resources are used by an application or more appropriately to context of this dissertation: \textit{what kind of resources can we add to improve application performance?} Required resources varies from an application to an application. However, among the most common ones we can distinguish:
\begin{itemize}
	\item CPU
	\item memory
	\item storage
	\item network bandwidth
\end{itemize}

@@

With all that said, there is no silver bullet - not matter what underlying mechanism platform provider decides to use, the application developer is still responsible for creating an application with scaling in-mind. This statement has been already proven in 1967 by Amdahl law, which in short states that sequential component of parallel algorithms impacts efficiency for a sufficiently large number processors \cite{Am67} as shown in Figure \ref{ch2:amdahl-law}. In other words, adding supplementary resources to a poorly written application (i.e. having a lot of sequential or synchronized components) can be beneficial only to a certain degree.

@@

\section{Horizontal scaling}
As outlined in previous section, horizontal scaling is about adding supplementary nodes to a system. As it is common to cloud computing, nodes are represented as virtual machines and this assumption is used in further discussion. Consequently, adding server comes down to cloning a new virtual machine from a template and possibly installing additional software and reconfiguring it later. While mechanism of creating new virtual machine from a template is offered literally in every IaaS platform currently available (OpenStack \cite{OpenStack}, OpenNebula \cite{OpenNebula}, CloudStack \cite{CloudStack} or Eucalyptus \cite{Eucalyptus} to name a few) and is similar in manner, the underlying hardware and virtualization mechanism determines how fast provisioning is done.

@@

\subsection{Load-balancing algorithms}
Generally, there are two types of load-balancers: hardware and software based. Due to the dynamic nature of system under consideration, we focus only on the latter as it offers a greater deal of flexibility. Among the most common algorithms we can distinguish \cite{HaProxyDoc}:
\begin{itemize}
 \item \textit{round-robin scheduling} - request are sent to sucessive nodes, according to their weights. This algorithm is fairest when the server's processing time remains equally distributed \cite{HaProxyDoc}
 \item \textit{least connection} - the server with the lowest number of connections receives the connection
 \item \textit{source routing} - source IP address is hashed, the same client IP address always reaches the same server
 \item \textit{URI hashing} - URI that designates resource is hashed and divided by the total weight of the running servers. Such hash designates which server that receives the request. In practice, this algorithm is commonly used with proxy caches and anti-virus proxies in order to maximize the cache hit rate.
\end{itemize}

@@

Below are presented one of the most common load-balancers, including their key performance features.

\subsubsection*{HAProxy}

HAProxy \cite{HAProxy} is a load-balancer initially written by Willy Tarreau. Noticeably, it's used by OpenShift \cite{OpenShift} to distribute load among gears \cite{OpenShiftScaling}. Its key features include:
\begin{itemize}
	\item a single-process, event-driven model reduces the cost of context switch and the memory usage
	\item O(1) event checker
	\item Single-buffering without copying data between reads and writes
	\item Zero-copy forwarding
	\item Optimized HTTP header analysis : headers are parsed an interpreted on the fly
\end{itemize}

Requests are dispatched accordingly to chosen algorithm:
\begin{itemize}
  \item \textit{round-robin scheduling}
  \item \textit{least connection}
  \item \textit{source routing}
  \item \textit{URI hashing}
\end{itemize}

\subsubsection*{BIG-IP Local Traffic Manager (LTM)}
BIG-IP Local Traffic Manager is a solution offered by F5 \cite{F5}. Although LTM is a hardware solution, omitted in this section, it also has its virtualized counterpart. Key features offered by LTM are \cite{LTM}:
\begin{itemize}
  \item managing at application services level rather than at individual devices and objects
  \item scripting language that allows administrator to intercept, inspect, transform, and direct application traffic
  \item built-in firewall protection, application security, and access control
  \item real-time protocol and traffic management decisions
\end{itemize}

\subsubsection*{Apache HTTPD}
Apache HTTPD \cite{ApacheHTTPD} is a popular HTTP server. When enhanced with additional modules, it can behave as a proxy or load-balancer. Over the time, there were serveral attempts to develop such modules: mod\_jk \cite{ApacheModJk}, mod\_proxy\_balancer \cite{ApacheModProxyBalancer}, to name a few. While the former is AJP13 oriented, the latter supports different protocols: HTTP, FTP and AJP13, hence, it is described here.

Among the mod\_proxy\_balancer features we may distinguish:
\begin{itemize}
 \item support for different load-balancing algorithms:
    \begin{itemize}
      \item \textit{request counting algorithm} - load is distributed the requests among the various workers, ensuring that each gets their configured share of the number of requests
      \item \textit{weighted traffic counting algorithm} - variation of above-mentioned algorithm with a difference that it is focused on bytes rather than number of request
      \item \textit{pending request counting algorithm} - scheduler keeps track of how many requests each worker is assigned at present. A new request is automatically assigned to the worker with the lowest number of active requests
    \end{itemize}

 \item support for session stickiness by using cookies and URL encoding. This approach \cite{ApacheModProxyBalancer} avoids unequal load distribution if clients are hidden behind proxies and stickyness errors when a client uses a dynamic IP address that changes during a session
\end{itemize}

@@

\section{Vertical scaling}
Essentially, vertical scaling is concentrated upon increasing capacity of single node. Again, when considering technical advancements that comes with cloud computing and virtualization, we can differ two categories of scaling: virtual machine resizing and virtual machines replacement. This distinction is dictated by limitation hypervisors - not all of them are able to resize virtual machine without shutting it down.

\subsection*{Virtual machine resizing}

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}{| l | X | X | X |}
\hline
 & \textbf{Memory} & \textbf{CPU} & \textbf{Disk} \\
\hline 
KVM &
-  & 
dynamic pinning CPU to a specific virtual machine (depending on underlying hardware)
& 
adding a disk to a LVM group

\\ \hline
Xen & 
- changing the amount of host physical memory assigned to virtual machine without rebooting it

- start additional virtual machines on a host whose physical memory is currently full, by automatically reducing the memory allocations of existing virtual machines in order to make space
&
dynamic pinning CPU to a specific virtual machine (depending on underlying hardware)
&
dynamic block attaching, adding a disk to a LVM group

\\ \hline
VMware ESX &
hot-plugging memory, ex. using VMware vSphere
&
hot-plugging CPU, ex. using VMware vSphere
&
adding additional disks to existing virtual machine

\\ \hline
OpenVZ &
configurable via user beancounters
&
configurable via user beancounters 
& 
configurable via user beancounters
\\ \hline
\end{tabularx}
\caption{Comparision of hypervisors resizing capabilities}
\label{tab:hypervisors-resizing}
\end{table}

\subsection*{Virtual machine replacement}
As it was highlighted in previous section, reasoning behind virtual machine replacement is that, in case when dynamic resizing is not possible, a new virtual machine with a desired configuration can be provisioned and replace the old one. Since this is a basic operation, all above-mentioned hypervisors supports this scenario as long as required resources are available.

@@

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}[]{ X  c  c  c }
\specialrule{.1em}{.05em}{.05em}

  & \textbf{Horizontal scaling} & \textbf{Vertical scaling} & \textbf{Application / resource tuning} \\
\specialrule{.1em}{.05em}{.05em}

>>>>

While scalability is a widely used term, it still lacks a clear and concise definition. Over the time, there were a few attempts to define it, yet not all of them were claimed as successful \cite{Hi90} \cite{DuRoWi06}. Hence, it is necessary to clarify this term before going into further discussion. Instinctively, scalability is perceived as ability of a system to accommodate an increasing number of elements or objects to process. In particular, we can point out different types of scalability that are affected by increased number of requests: \cite{Bo00}:
\begin{itemize}
	\item \textit{load scalability} - ability to work without delays and unproductive resource consumption at light, moderate, or heavy loads while making good use of available resources. Factors that may hinder load scalability include: scheduling shared resource, self-expansion, inadequate exploitation of parallelism
	\item \textit{space scalability} - memory requirements do not grow to intolerable levels as the number of items system supports increases
	\item \textit{space-time scalability} - system continues to function gracefully as the number of objects it encompasses increases by orders of magnitude
	\item \textit{structural scalability} - implementation or standards do not impede the growth of the number of objects system encompasses
\end{itemize}
Although, all of the aforementioned aspects are vital for any application, our work focuses solely on the first type of scalability. The reasoning behind this statement is that, while all of these scalability types lies in direct responsibility of an application developer, the load scalability can be additionally improved by adding additional resources to a system. This brings us to a question what kind of resources are used by an application or more appropriately in context of this dissertation: \textit{what kind of resources can we add to improve application performance?} Required resources varies from an application to an application. However, among the most common ones we can distinguish:
\begin{itemize}
	\item CPU
	\item memory
	\item storage
	\item network bandwidth
\end{itemize}

@@

With all that said, there is no silver bullet - not matter what underlying mechanism platform provider decides to use, the application developer is still responsible for creating an application with scaling in-mind. This statement has been already proven in 1967 by Amdahl law, which in short states that sequential component of a parallel algorithm impacts efficiency for a sufficiently large number processors \cite{Am67} as shown in Figure \ref{ch2:amdahl-law}. In other words, adding supplementary resources to a poorly written application (i.e. having a lot of sequential or synchronized components) can be beneficial only to a certain degree.

@@

\newpage
\section{Horizontal scaling}
As outlined in previous section, horizontal scaling is about adding supplementary nodes to a system. As it is common to cloud computing, nodes are represented as virtual machines and this assumption is used in further discussion. Consequently, adding server comes down to cloning a new virtual machine from a template and possibly installing additional software and reconfiguring it later. While mechanism of creating new virtual machine from a template is offered literally in every IaaS platform currently available (OpenStack \cite{OpenStack}, OpenNebula \cite{OpenNebula}, CloudStack \cite{CloudStack} or Eucalyptus \cite{Eucalyptus} to name a few) and is similar in manner, the underlying hardware and virtualization mechanism determines how fast provisioning is done.

@@

\subsection{Load-balancing algorithms}
Generally, there are two types of load-balancers: hardware and software based. Due to the dynamic nature of system under consideration, we focus only on the latter as it offers a greater deal of flexibility. Among the most common algorithms we can distinguish \cite{HaProxyDoc}:
\begin{itemize}
 \item \textit{round-robin scheduling} - request are sent to sucessive nodes, according to their weights. This algorithm is fairest when the server's processing time remains equally distributed \cite{HaProxyDoc}
 \item \textit{least connection} - the server with the lowest number of connections receives the connection
 \item \textit{source routing} - source IP address is hashed, the same client IP address always reaches the same server
 \item \textit{URI hashing} - URI that designates resource is hashed and divided by the total weight of the running servers. Such hash designates which server that receives the request. In practice, this algorithm is commonly used with proxy caches and anti-virus proxies in order to maximize the cache hit rate.
 \item \textit{request counting algorithm} - load is distributed the requests among the various workers, ensuring that each gets their configured share of the number of requests
 \item \textit{weighted traffic counting algorithm} - variation of above-mentioned algorithm with a difference that it is focused on bytes rather than number of request
 \item \textit{pending request counting algorithm} - scheduler keeps track of how many requests each worker is assigned at present. A new request is automatically assigned to the worker with the lowest number of active requests
\end{itemize}

@@

@@

\subsection{Load-balancer comparision}

While there are many load-balancers available on the market, following are credited to be most popular:
\begin{itemize}
 \item \textbf{HAProxy} \cite{HAProxy} - load-balancer initially written by Willy Tarreau. Noticeably, it's used by OpenShift \cite{OpenShift} to distribute load among gears \cite{OpenShiftScaling}
 \item \textbf{BIG-IP Local Traffic Manager (LTM)} - solution offered by F5 \cite{F5}. Although LTM is a hardware solution, omitted in this section, it also has also its virtualized counterpart.
 \item \textbf{Apache HTTPD} \cite{ApacheHTTPD} - popular HTTP server. When enhanced with additional modules, it can behave like a proxy or load-balancer. Over the time, there were serveral attempts to develop such modules: mod\_jk \cite{ApacheModJk}, mod\_proxy\_balancer \cite{ApacheModProxyBalancer}, to name a few. While the former is purely AJP13 oriented, the latter supports different protocols: HTTP, FTP and AJP13. As a consequnce, only mod\_proxy\_balancer was taken into account during comparision.
\end{itemize}

Table \ref{tab:load-balancer-comparision} presents they key performance features and algorithm used to schedule requests.

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}[]{ X  X  X }
\specialrule{.1em}{.05em}{.05em}

  & \textbf{Performance features} & \textbf{Scheduling algorithms} \\
\specialrule{.1em}{.05em}{.05em}

HAProxy & 
-- a single-process, event-driven model reduces the cost of context switch and the memory usage

-- O(1) event checker

-- single-buffering without copying data between reads and writes

-- zero-copy forwarding

-- optimized HTTP header analysis: headers are parsed an interpreted on the fly

&
-- round-robin scheduling

-- least connection

-- source routing

-- URI hashing
\\ \hline

BIG-IP Local Traffic Manager &
-- managing at application services level rather than at individual devices and objects

-- scripting language that allows administrator to intercept, inspect, transform, and direct application traffic

-- built-in firewall protection, application security, and access control

-- real-time protocol and traffic management decisions
&

\\ \hline

Apache HTTPD & 
-- support for session stickiness by using cookies and URL encoding. This approach \cite{ApacheModProxyBalancer} avoids unequal load distribution if clients are hidden behind proxies and stickyness errors when a client uses a dynamic IP address that changes during a session
& 
-- request counting algorithm

-- weighted traffic counting algorithm

-- pending request counting algorithm
\\ \hline

\end{tabularx}

\caption{Comparision of load balancers}
\label{tab:load-balancer-comparision}

\newpage
\section{Vertical scaling}
Essentially, vertical scaling is concentrated upon increasing capacity of single node. Again, when considering technical advancements that comes with cloud computing and virtualization, we can differ two categories of scaling: virtual machine resizing and virtual machines replacement. This distinction is dictated by limitation hypervisors - not all of them are able to resize virtual machine without shutting it down.

\subsection{Virtual machine resizing}

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}{ l  X  X  X }
\specialrule{.1em}{.05em}{.05em} 
 & \textbf{Memory} & \textbf{CPU} & \textbf{Disk} \\
\specialrule{.1em}{.05em}{.05em}

KVM &
  & 
-- dynamic pinning CPU to a specific virtual machine (depending on underlying hardware)
& 
-- adding a disk to a LVM group

\\ \hline
Xen & 
-- changing the amount of host physical memory assigned to virtual machine without rebooting it

-- start additional virtual machines on a host whose physical memory is currently full, by automatically reducing the memory allocations of existing virtual machines in order to make space
&
-- dynamic pinning CPU to a specific virtual machine (depending on underlying hardware)
&
-- dynamic block attaching, adding a disk to a LVM group

\\ \hline
VMware ESX &
-- hot-plugging memory, ex. using VMware vSphere
&
-- hot-plugging CPU, ex. using VMware vSphere
&
-- adding additional disks to existing virtual machine

\\ \hline
OpenVZ &
-- configurable via user beancounters
&
-- configurable via user beancounters 
& 
-- configurable via user beancounters
\\ \hline
\end{tabularx}
\caption{Comparision of hypervisors resizing capabilities}
\label{tab:hypervisors-resizing}
\end{table}

\subsection{Virtual machine replacement}
As it was highlighted in previous section, reasoning behind virtual machine replacement is that, in case when dynamic resizing is not possible, a new virtual machine with a desired configuration can be provisioned and replace the old one. Since this is a basic operation, all above-mentioned hypervisors supports this scenario as long as required resources are available.

@@

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}[]{ X  X  X  X}
\specialrule{.1em}{.05em}{.05em}

  & \textbf{Horizontal scaling} & \textbf{Vertical scaling} & \textbf{Application tuning} \\
\specialrule{.1em}{.05em}{.05em}

### 2013-09-13 thesis/src/chapters/chapter-3.tex
refactored table

>>>>

Table \ref{tab:cloud-providers-adaptivity} summaries chapter with approaches to adaptivity taken by different cloud providers. Section 'triggered action' is omited for brevity - it was described extensively in previous chapter.

\begin{table}[!htbp]
\begin{tabularx}{\textwidth}[]{ X  X X }
\specialrule{.1em}{.05em}{.05em}

  & \textbf{Policies} & \textbf{Data analysis} \\
\specialrule{.1em}{.05em}{.05em}

\multicolumn{3}{ l }{\textbf{Infrastructure provider}} \\
\specialrule{.1em}{.05em}{.05em}

Carina & 
- time frame based

- expression based (only for CPU)
&
- threshold model

\\ \hline

OneFlow & 
- time frame based

- expression based
&
- threshold model
\\ \hline

AWS EC2 & 
- time frame based

- expression based, (CloudWatch alarm)
&
- threshold model
\\ \hline

\multicolumn{3}{ l }{\textbf{Platform provider}} \\
\specialrule{.1em}{.05em}{.05em}

CloudFoundry & $\times$ & $\times$ \\ \hline

OpenShift & 
$\times$ &
single built-in threshold model 
\\ \hline

AppEngine & ? & ? \\ \hline

Azure & ? & ? \\ \hline

Heroku & ? & ? \\ \hline
\end{tabularx}

\caption{Comparision of cloud providers approach to adaptivity}
\label{tab:cloud-providers-adaptivity}

\end{table}
